Synthesis of Linear Functional Programs

### 2021-07-20 article/article.tex
tweaks

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a
system of resource management and focusing.
%
In intuitionistic sequent calculi, each connective has a so-called
\emph{left} and a \emph{right} rule, which effectively define how to
decompose an ambient assumption of a given proposition and how to
prove a certain proposition is true, respectively. 
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible
and non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Synchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Asynchronous connectives are dual. \todo{Say which
  are which}
%
Given this separation, focusing divides proof search into two
phases: %
the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
to focus on, and then apply non-invertible rules, staying \emph{in focus}
until we reach an asynchronous, i.e. invertible,
proposition or the proof is complete.

@@

The \synname\ synthetiser operates as part of the pipeline that processes a
full program in the \synname\ language, and is called to generate parts marked
(by the syntax \emph{\{\{ ... \}\}}) for
synthesis of said program. The main pipeline consists of:
\[
    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow \textrm{Inference} \rightarrow \textrm{Synthesis}
\]

@@

The desugaring module converts the frontend AST to a core AST without syntatic
sugar and using the locally-nameless representation~\cite{locally nameless}, in
which all bound variables have no name, and instead are addressed using the
number of binders traversed since the binding.
% TODO: não ficou muito mais claro :\todo{Explica o que e isto}
This way inference can be done without worrying about name conflicts.

@@

The synthesis module iterates over the synth marks and runs the synthetiser with
its context and goal for each, minimizing the resulting expression, and
replacing the mark in the frontend AST with the synthed program.

@@

\mypara{General} The complete implementation of the synthetiser and environment
pipeline was done in Haskell...

\mypara{Backtracking} Even though the synthetiser formal system helps, there's
still non-determinism in the synthesis process. At diverse times in the proof
construction we must choose one of multiple rules to apply, and in the event of
not being able to construct the proof following that choice, we want to
backtrack to the decision point and attempt to create a proof taking a different
route. To this effect, our prototype synthetiser depends on \emph{logict}, "a backtracking logic-programming
monad"~\cite{logict}.

\mypara{SMT Solving} To typecheck and synthetise refinement types and dependent
functions, we need to contact an SMT solver to check satisfiability. For this,
we use the library \emph{sbv}~\cite{sbv}. Unfortunately, this library does not
support the usage of the SMT solvers to satisfy and output more complicated
first-order logic uninterpreted functions -- since this is needed to synthetise refinement types
when right focused, we instead use unsafe library primitives to communicate with
the solver directly, constructing the logical formulas ourselves. For 
satisfying formulas, for instance, when typechecking, or left focused on a
refinement type, we use the library api successfully to get a veridct on
satisfiability.

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
quite noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications. For example, if we add a new function
\emph{read} to the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take around
$5s$, while synthesis without memoization doesn't terminate even after
$15m$ running.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some examples.

>>>>

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a
system of resource management and focusing.
%
In intuitionistic sequent calculi, each connective has a so-called
\emph{left} and a \emph{right} rule, which effectively define how to
decompose an ambient assumption of a given proposition and how to
prove a certain proposition is true, respectively. 
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible
and non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Synchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Asynchronous connectives are dual. \todo{Say which
  are which}
%
Given this separation, focusing divides proof search into two
phases: %
the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
to focus on, and then apply non-invertible rules, staying \emph{in focus}
until we reach an asynchronous (i.e. invertible proposition), the
proof is complete or no rules are applicable, in which the proof must
\emph{backtrack} to the state at which the focusing phase began.

@@

The \synname\ synthetiser operates as part of the pipeline that processes a
full program in the \synname\ language, and is called to generate terms marked
(by the syntax \emph{\{\{ ... \}\}}) for
synthesis. The main pipeline consists of:
\[
    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow \textrm{Inference} \rightarrow \textrm{Synthesis}
\]

@@

The desugaring module converts the frontend AST to a core AST without
syntatic sugar and using the locally-nameless
representation~\cite{locally nameless}, in which all bound variables
are represented with Debruijn indices rather than names (i.e.~using
the distance to the binding site in terms of the number of traversed
binders),
% TODO: não ficou muito mais claro :\todo{Explica o que e isto}
This way inference can be done without worrying about name conflicts.

@@

The synthesis module iterates over the synthesis marks and runs the
synthetiser with its context and goal for each mark, minimizing the
resulting expression (i.e.~$\eta$-reducing it), and replacing the mark
in the frontend AST with the synthed program.

@@

\mypara{General} The implementation was carried out in Haskell. We
highlight some key points.

\mypara{Backtracking} Even though focusing reduces non-determinism
during proof search, it cannot fully eliminate it. At several points
during proof construction we must choose one of multiple rules to
apply, and in the event of not being able to construct the proof
following that choice, we want to backtrack to the decision point and
attempt to derive a proof taking a different route. To this effect,
our prototype synthetiser depends on \emph{LogicT}, "a backtracking
logic-programming monad"~\cite{logict}.

\mypara{SMT Solving} To typecheck and synthesize refinement types and
dependent functions, we interface with an SMT solver to check
satisfiability. For this, we use the library
\emph{SBV}~\cite{sbv}. Unfortunately, this library does not support
the usage of SMT solvers to check first-order logic with uninterpreted
functions. Since this is needed for synthesis with refinement types
when right focused, we instead use unsafe library primitives to
communicate with the solver directly, constructing the logical
formulas directly. For satisfying formulas, for instance, when
typechecking, or left focused on a refinement type, we use the library
API as an oracle.

\mypara{Memoization} Some proofs attempt to synthetise the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthetise result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values and
uses the result as the key\todo{via a hashconsing library?}.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We capture failure of a LogicT computation using the $\mathit{ifte}$
function and represent failure or success information with the \texttt{Maybe}
type. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take
around $5s$, while synthesis without memoization doesn't terminate
even after $15m$ running.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some
examples. We leave such optimizations to future work.
